### 2015-05-10 swmpr_manu.Rnw
all acronyms defined using glossaries package, added glossary to beginning of text

% Template for PLoS
% Version 3.1 February 2015
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% % % % % % % % % % % % % % % % % % % % % % %

@@

% packages I added
\usepackage{cleveref}
\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}

@@

@@

% acronyms
\acrodef{CDMO}{Centralized Data Management Office}
\acrodef{CRAN}{Comprehensive R Archive Network}
\acrodef{DO}{dissolved oxygen}
\acrodef{NERRS}{National Estuarine Research Reserve System}
\acrodef{QAQC}{quality assurance/quality control}
\acrodef{SWMP}{System Wide Monitoring Program}

@@

\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example in the United States is the \ac{SWMP} that was implemented in 1995 by the federally-funded \ac{NERRS}.  This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.  \ac{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with \ac{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.

\linenumbers

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acl{NERRS} (\acs{NERRS}, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acf{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \ac{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \ac{SWMP} data.  \ac{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \ac{SWMP} \cite{SWMP14}.

This article describes a software package that was developed to address research needs of the \ac{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.  SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the \acl{SWMP}.  Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.  Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.  The software is provided specifically for use with \ac{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.

@@

Four core data elements are collected through the \ac{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \ac{SWMP} data and, consequently, the most challenging to evaluate.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \ac{CDMO} web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple \ac{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \ac{QAQC} flags with the appropriate qualifier.

The \ac{CDMO} web portal was established to support priority areas of \ac{SWMP} that focus on the continuation and advancement of data management.  As such, \ac{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \ac{QAQC} procedures.    Prior to any data request to the \ac{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the \ac{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the \ac{CDMO} website. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered with \ac{CDMO} before using the data retrieval functions in SWMPr.  Web services are provided by \ac{CDMO} for direct access to \ac{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \ac{CDMO} web services.  For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \ac{SWMP} sites.  The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.

@@

The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be downloaded from the \ac{CRAN} by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

@@

Two approaches can be used to import \ac{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \ac{CDMO} web services. To do so, the IP address for the computer making the request must be registered by following instructions on the \ac{CDMO} website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \ac{CDMO}.

@@

descrips <- c(
  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\ac{CDMO} using the zip downloads option.',
  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
  )

@@

Data retrieval functions to import data directly into R from the \ac{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested. Due to rate limitations on the \ac{CDMO} server, the retrieval functions return a limited number of records with each request.  However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain the desired time series.  Download time can be excessive for longer time series.

@@

The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \ac{CDMO}.  This approach is most appropriate for large, specific data requests. The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the \ac{CDMO} website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

@@

All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.  The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).  Correspondingg \ac{QAQC} columns for each parameter are also returned if provided by the initial data request.  The following shows an example of the raw data imported using \texttt{all\_params}.

@@

# view all attributes of dat
atts <- paste0('\\texttt{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class',
  'Row names of the data set, inherited from the \\texttt{data.frame} object class',
  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}',
  'Station identifier used by \\ac{NERRS} as a string with 7 or 8 characters',
  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}",
  'Indicates if \\ac{QAQC} columns are present in the raw data',
  'Start and end dates for the raw data',
'Timezone of the station using the city/country format\\textsuperscript{a}',
  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated'
)

@@

A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).  This dataset has an identical format as the data returned from the zip downloads feature of the \ac{CDMO}.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the \ac{QAQC} flag that was assigned by \ac{CDMO} during initial processing.  The \ac{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \ac{QAQC} flag (i.e., passed initial \ac{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

descrips <- c(
  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove \\ac{QAQC} columns and remove data based on \\ac{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\ac{QAQC} columns are present. ', 
  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\ac{QAQC} flags used by \\ac{CDMO}.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \\texttt{comb} and can also be used with individual stations.',
  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
  )

@@

Viewing the number of observations for each parameter that are assigned to a \ac{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.

@@

Raw nutrient data obtained from the \ac{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \ac{QAQC} columns are present.

@@

Detailed exploratory graphics are also useful for evaluating general trends in observed data.  Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (\Cref{fig:summary_ex}).  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.  An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \ac{SWMP} sites (see the \nameref{swmp_apps} section).

@@

Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}. The diffusion-corrected \ac{DO} flux estimates are averaged separately over each day and night of the time series. The nighttime average \ac{DO} flux is used to estimate respiration rates, while the daytime \ac{DO} flux is used to estimate net primary production. To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \ac{DO} flux rates are multiplied by 24. Similarly, the daytime \ac{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production. Respiration rates are subtracted from daily net production estimates to yield gross production rates. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the \ac{DO} time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

@@

Several additional functions are provided that do not fit the above categories (\Cref{tab:misc}).  These functions are generally used within the main functions but may be useful for more customized evaluation of \ac{SWMP} data.

@@

cap <- 'Trends in metabolism for two sites at each of the \\ac{NERRS} reserves.  Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014). Bold values indicate an increase from the first period, whereas italic values indicate a decrease. Sites were assigned to regions based on approximate geographic coordinates.' 
foots <- '\\footnotesize{\\textsuperscript{{\\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'

>>>>

% Template for PLoS
% Version 3.1 February 2015

@@

@@

% packages I added
\usepackage{cleveref}
\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[acronym]{glossaries}
\makeglossaries
\include{glossary}

@@

% acronyms from glossaries package
\newacronym{CDMO}{CDMO}{Centralized Data Management Office}
\newacronym{CRAN}{CRAN}{Comprehensive R Archive Network}
\newacronym{DO}{DO}{dissolved oxygen}
\newacronym{NERRS}{NERRS}{National Estuarine Research Reserve System}
\newacronym{QAQC}{QAQC}{quality assurance/quality control}
\newacronym{SWMP}{SWMP}{System Wide Monitoring Program}

@@

\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \gls{NERRS}.  This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.  \gls{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with \gls{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
\linenumbers
\glsaddall
\printglossaries

@@

An invaluable source of monitoring data for coastal environments in the United States is provided by the \acrlong{NERRS} (\acrshort{NERRS}, \url{http://www.nerrs.noaa.gov/}).  This network represents 28 estuarine reserves from different biogeographic regions that were chosen to address multiple goals for long-term research, monitoring, education, and stewardship in support of coastal management.  As part of this effort, the \acrlong{SWMP} was implemented in 1995 at over 140 stations across the reserves to provide a robust, long-term monitoring system for water quality, weather, and land-use/habitat change.  The \gls{SWMP} network has provided a continuous source of data collected at near real-time with the intent to evaluate natural and anthropogenic causes of spatiotemporal variation in environmental condition and ecosystem function.  These data have been applied both for evaluations of relevant characteristics at individual reserves (eg., \cite{Bulthius95,Dix08}) and differences between reserves (e.g., ecosystem metabolism \cite{Caffrey03,Caffrey04}, tidal characteristics \cite{Sanger02}, dissolved oxygen \cite{Wenner04}).  However, no cross-reserve comparisons have been conducted within the last decade despite the online availability of current \gls{SWMP} data.  \gls{NERRS} researchers and staff have also expressed a need for quantitative analysis tools to evaluate trends in water quality time series given the quantity and quality of data provided by \gls{SWMP} \cite{SWMP14}.

This article describes a software package that was developed to address research needs of the \gls{NERRS} program using the open-source statistical programming language R \cite{RDCT14}.  SWMPr (pronounced `swamper') is an R package that contains functions for retrieving, organizing, and analyzing estuary monitoring data from the \acrlong{SWMP}.  Functions provided by SWMPr address many of the common issues working with large datasets created from automated sensor networks, such as data pre-processing to remove unwanted information, combining data from different sources, and exploratory analyses to identify key parameters of interest.  Additionally, a cross-reserve comparison of water quality trends and current ecosystem metabolism estimates is provided to illustrate potential applications using the functions in this package.  The software is provided specifically for use with \gls{NERRS} data, although many of the applications are relevant for addressing common challenges working with large datasets.

@@

Four core data elements are collected through the \gls{SWMP} monitoring network: abiotic monitoring data, biotic observations, habitat and land use mapping, and sentinel monitoring.  Only the abiotic data are monitored continuously with automated sensor networks, whereas the remaining elements involve field surveys or mapping products that differ between reserves given site-specific requirements.  As such, the SWMPr package is developed for the continuous abiotic monitoring network that represents a majority of the \gls{SWMP} data and, consequently, the most challenging to evaluate.  Abiotic elements monitored at each reserve include water quality (water temperature, specific conductivity, salinity, dissolved oxygen concentration, dissolved oxygen saturation, depth, pH, turbidity, chlorophyll fluorescence), weather (air temperature, relative humidity, barometric pressure, wind speed, wind direction, photosynthetically active radiation, precipitation), and nutrient data (orthophosphate, ammonium, nitrite, nitrate, nitrite + nitrate, chlorophyll a).  Each reserve has no less than four water quality stations and one weather station at fixed locations.  Water quality and weather data are collected at 15 minute intervals, whereas nutrient data are collected monthly at each water quality station.  All data are made accessible through the \gls{CDMO} web portal (\url{http://cdmo.baruch.sc.edu/}), where multiple \gls{QAQC} measures are used to screen the information for accuracy and reliability.  The final data include all timestamped observations including relevant \gls{QAQC} flags with the appropriate qualifier.

The \gls{CDMO} web portal was established to support priority areas of \gls{SWMP} that focus on the continuation and advancement of data management.  As such, \gls{CDMO} provides access to over 35 million water quality, weather, and nutrient records that have been authenticated through systematic \gls{QAQC} procedures.    Prior to any data request to the \gls{CDMO}, the location, parameter type, and date ranges need to be identified based on the analysis needs. All stations in the \gls{SWMP} network are identified by a 7 or 8 character name that specifies the reserve, station, and parameter type.  For example, `apaebwq' is the water quality identifier (`wq') for the East Bay station (`eb') at the Apalachicola reserve (`apa').  Similarly, a suffix of `met' or `nut' would specify the weather (meteorological) or nutrients station.  All reserve names, stations, and date ranges for each parameter type can be viewed on the \gls{CDMO} website. Alternatively, the \texttt{site\_codes} (all sites) or \texttt{site\_codes\_ind} (single site) functions provided by the SWMPr package can be used to view the same information.  As noted below, the computer's IP address must be registered with \gls{CDMO} before using the data retrieval functions in SWMPr.  Web services are provided by \gls{CDMO} for direct access to \gls{SWMP} data through http requests, in addition to standard graphical user interface options for selecting data.  The data retrieval functions in SWMPr are simple calls to the existing retrieval functions on \gls{CDMO} web services.  For example, the \texttt{site\_codes} function in SWMPr uses the \texttt{exportStationCodesXMLNew} function from the web services to retrieve metadata for all the \gls{SWMP} sites.  The text below describes the data retrieval functions in more detail, including all other functions available in SWMPr.

@@

The SWMPr package was developed for use with the R ($\geq$ v3.0.0) statistical programming language \cite{RDCT14}. The SWMPr package can be downloaded from the \gls{CRAN} by executing the following commands at the R console prompt.  The package is loaded in the workspace using the \texttt{library} command.

@@

Two approaches can be used to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \gls{CDMO} web services. To do so, the IP address for the computer making the request must be registered by following instructions on the \gls{CDMO} website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \gls{CDMO}.

@@

descrips <- c(
  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\gls{CDMO} using the zip downloads option.',
  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
  )

@@

Data retrieval functions to import data directly into R from the \gls{CDMO} include \texttt{all\_params}, \texttt{all\_params\_dtrng}, and \texttt{single\_param}: \texttt{all\_params} returns the most recent records of all parameters at a station, \texttt{all\_params\_dtrng} returns all records within a date range for all parameters or a single parameter, and \texttt{single\_param} is identical to \texttt{all\_params} except that a single parameter is requested. Due to rate limitations on the \gls{CDMO} server, the retrieval functions return a limited number of records with each request.  However, the SWMPr functions use the native CDMO web services iteratively (i.e., within a loop) to obtain the desired time series.  Download time can be excessive for longer time series.

@@

The second approach for data retrieval is to use the \texttt{import\_local} function to import data into R that are locally available after downloading from \gls{CDMO}.  This approach is most appropriate for large, specific data requests. The \texttt{import\_local} function is designed for data from the zip downloads feature in the advanced query section of the \gls{CDMO} website. The zip downloads feature can be used to obtain data from multiple stations in one request.  The downloaded data will be in a compressed folder that includes multiple .csv files by year for a given data type (e.g., apacpwq2002.csv, apacpwq2003.csv, apacpnut2002.csv, etc.).  The \texttt{import\_local} function can be used to import files directly from the zipped folder or after the folder is decompressed.

@@

All data retrieval functions return a \texttt{swmpr} object that includes relevant data and several attributes describing the dataset.  The data include a \texttt{datetimestamp} column in the appropriate timezone for a station and additional parameters for a given data type (weather, nutrients, or water quality).  Correspondingg \gls{QAQC} columns for each parameter are also returned if provided by the initial data request.  The following shows an example of the raw data imported using \texttt{all\_params}.

@@

# view all attributes of dat
atts <- paste0('\\texttt{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class',
  'Row names of the data set, inherited from the \\texttt{data.frame} object class',
  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}',
  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters',
  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}",
  'Indicates if \\gls{QAQC} columns are present in the raw data',
  'Start and end dates for the raw data',
'Timezone of the station using the city/country format\\textsuperscript{a}',
  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated'
)

@@

A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).  This dataset has an identical format as the data returned from the zip downloads feature of the \gls{CDMO}.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.  The \gls{QAQC} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

descrips <- c(
  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove \\gls{QAQC} columns and remove data based on \\gls{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\gls{QAQC} columns are present. ', 
  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to different \\gls{QAQC} flags used by \\gls{CDMO}.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is used in \\texttt{comb} and can also be used with individual stations.',
  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
  )

@@

Viewing the number of observations for each parameter that are assigned to a \gls{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.

@@

Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA}, values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.

@@

Detailed exploratory graphics are also useful for evaluating general trends in observed data.  Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (\Cref{fig:summary_ex}).  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.  An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \gls{SWMP} sites (see the \nameref{swmp_apps} section).

@@

Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}. The diffusion-corrected \gls{DO} flux estimates are averaged separately over each day and night of the time series. The nighttime average \gls{DO} flux is used to estimate respiration rates, while the daytime \gls{DO} flux is used to estimate net primary production. To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \gls{DO} flux rates are multiplied by 24. Similarly, the daytime \gls{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production. Respiration rates are subtracted from daily net production estimates to yield gross production rates. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the \gls{DO} time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a significant amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

@@

Several additional functions are provided that do not fit the above categories (\Cref{tab:misc}).  These functions are generally used within the main functions but may be useful for more customized evaluation of \gls{SWMP} data.

@@

cap <- 'Trends in metabolism for two sites at each of the \\gls{NERRS} reserves.  Values are averages of mean annual estimates for each period of observation (1994-2004 and 2005-2014). Bold values indicate an increase from the first period, whereas italic values indicate a decrease. Sites were assigned to regions based on approximate geographic coordinates.' 
foots <- '\\footnotesize{\\textsuperscript{{\\it a}}NEM: net ecosystem metabolism, Pg: gross production, Rt: total respiration, all values in g O$_2$ m$^{-2}$ d$^{-1}$ as annual averages.}'
